Learning a reward function from demonstrations: a cascaded supervised learning approach

### 2012-10-26 piot2012cascading/ECML13_cascading.tex
version 2 du draft

% Created 2012-10-15 lun. 14:49
\documentclass[smallextended]{svjour3}
\smartqed 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\tolerance=1000
\usepackage{amsmath}
%\usepackage{amsthm}
%\newtheorem{definition}{Definition}
%\newtheorem{theorem}{Theorem}
\usepackage{dsfont}
\providecommand{\alert}[1]{\textbf{#1}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\begin{document}
\title{Learning a reward function from demonstrations: a cascaded supervised learning approach}
\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
\institute{
 1. ABC Team\\
 LORIA-CNRS, France.\\
\and
2.Supélec-Metz Campus\\
 MaLIS Research group, France\\
 \texttt{firstname.lastname@supelec.fr}\\
\and 
3. UMI 2958 CNRS\\
GeorgiaTech, France\\
}
\date{\today}

@@

\begin{abstract}
  This paper considers the Inverse Reinforcement Learning (IRL) problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
We propose to break the IRL problem down into two Supervised Learning (SL) steps: this is the cascaded approach. A classification step aiming to imitate the expert's policy gives a score function we see as an optimal action value function. In the second step, this score function is used together with the bellman evaluation equation to create samples then fed to a regressor that outputs a reward function.
This cascaded approach is justified by an analysis which shows that the demonstrated expert policy is near-optimal for the reward function the regression step outputs. We explain some advantages of this approach over existing algorithms, mainly not needing to repeatedly solve the Reinforcement Learning (RL) direct problem.
Finally, up to the use some heuristics, we empirically show the algorithm to work with only transitions sampled according to the demonstrated expert policy.
  \end{abstract}
\section{Introduction}
\label{sec-2}

Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} aims at inferring a reward function for which a demonstrated expert policy is optimal. An important part of the literature (see Section~\ref{sec:related} for a brief review) tries to find a reward function such that the associated optimal policy induces a distribution over trajectories (or some measure of this distribution) which matches the one induced by the expert: this is called apprenticeship learning. Most algorithms of apprenticeship learning use Reinforcement
Learning (RL) algorithms as subroutines, thus their performance depends strongly on the quality of these subroutines. Consequently, apprenticeship learning algorithms suffer from the
same challenges of large state spaces, exploration vs. exploitation trade-offs as RL. In order to avoid to solve repeatedly RL problems, we decide to take a different point of view.

Having in mind that there is a one to one relation between a reward function and its associated optimal quality function (via the inverse Bellman equation), it is worth thinking of a method which is able to output a quality function for which the greedy policy is the demonstrated expert policy. Thus, the demonstrated expert policy will be optimal for the reward function associated to the quality function. As a method which is able to output a quality function for which the greedy policy is the demonstrated expert policy, we propose a score-based classification (see Section~\ref{sec:algo}). Besides, in order to retrieve, via the inverse Bellman equation, the reward associated to the score function the classification outputs, we propose a regression (see Section~\ref{sec:algo}). That is why the method is called the Cascaded Supervised Approach (CSA). This method is justified by a theoretical analysis (see Section~\ref{sec:analysis}) where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.

Finally, an instantiation of CSA is proposed~\ref{sec:experiments}. This algorithm does not need to solve iteratively an RL problem and needs only sampled transitions from expert and non-experts policies as inputs. Moreover, up to the use of some heuristics, the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy. The algorithm is tested on a car driving simulator and compared to an apprenticeship learning algorithm, to a pure classification algorithm and to a recent successfull IRL method.

@@

We frame the sequential decision problem in a finite \emph{Markov Decision Process} (MDP) \cite{puterman1994markov}, that is a tuple $\{S,A,P,R,\gamma\}$. $S$ is a finite state space, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the dynamic of the MDP. Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$. At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one deterministic optimal policy.}) \emph{policy} $\pi\in A^S$. The agent then transits to a new state $s_{t+1}\in S$ according to the markovian transition probabilities $P(s_{t+1}|s_t,a_t)$. The reward function $R$ is a local measure of the quality of the control. The global quality of the control induced by a policy $\pi$, with respect to a reward $R$ is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates each state with the discounted cumulative reward for following policy $\pi$ from this state: $V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t)|s_0 = s]$. Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, \forall s \in S, V^*_R(s) \geq V^\pi_R(s)$. The stationary distribution induced by a policy $\pi$ is written $\rho_\pi,\textrm{ s.t. }\rho_\pi^TP^\pi = \rho_\pi$.

The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by $T^{\pi_R}V = R + \gamma P_\pi V$ where $P_\pi$ is $(p(s'|s,\pi(s)))_{s,s' \in S}$. The Bellman optimality operator follows naturally : $T^*_RV = \max_\pi T^\pi_RV$. Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R: $V^\pi_R$; the fixed point of the Bellman optimality operator $T^*_R$ is the optimal value function $V_R^*$ with respect to reward $R$. Optimal value function and optimal policy are also linked via the quality function $Q^\pi_R\in\mathbb{R}^{S\times A}$, a state-action value function that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^aV = R + \gamma P_a V$ with $P_a = (p(s'|s,a))_{s,s' \in S}$). The optimal policy follows a greedy mechanism with respect to its optimal quality function 
\begin{equation}
  \label{eq:greedy}
  \pi^*_R(s)\in\argmax_aQ^*_R(s,a).
\end{equation}

When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, RL or approximate dynamic programming may be used to retrieve the optimal control policy \cite{sutton1998reinforcement}. We call this the direct problem.

In this work, we wish to tackle the inverse problem. We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi^E$, assuming that there exist some unknown reward $R^E$ for which the expert is optimal. The non optimality of the expert is an interesting setting that has been discussed for example in \cite{melo2012analysis} but that we are not addressing here. We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is also near-optimal. The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal). Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see section \ref{sec:related}.

@@

Our first step in the quest for a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (MC$^2$). This classifier learn a score function $q:S\times A \rightarrow \mathbb{R}$ that rates the association of a certain label $a\in A$ with a certain input $s\in S$. The classification rule $\pi_C: S \rightarrow A$ is simply to select (one of) the label(s) that achieve the highest score for the given input :

@@

Most notably, \emph{Multi-class Support Vector Machines} can be seen as MC$^2$ algorithms. We give more examples and detail our choice of a MC$^2$ algorithms for the experiments in subsection \ref{subsec:instanciation}.

Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier. Apart from discussion and comparison purpose, we do not make use of the classification policy $\pi_C$. What is of interest to us is the score function $q$ itself. One can easily notice the similarity between equations \ref{eq:greedy2} and \ref{eq:greedy} that describes the relation between the optimal policy in a MDP and its optimal quality function. As the expert is optimal with respect to some unknown reward function $R_E$ and we seek a reward function $R$ for which it is also optimal, it is only natural to view the score function $q$ of the classifier as some kind of optimal quality function for the classifier policy $\pi_C$. By inversing the Bellman evaluation equation, one can express the reward as a function of the quality function : $R(s,a) = Q^*_R(s,a) - \sum_{s'}P(s'|s,a)Q^*_R(s'\pi^*(s'))$. This leave us with $r$, the reward function relative to our score/quality function $q$ : $r(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s'))$.

We wish to solve the general, or approximate, IRL problem where the transition probabilities are unknown. For this we use another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitioned to. Action $a_j$ need not be chosen by any particular policy. The dataset $D_R$ brings us information about the dynamics of the system. Our reward function $r$ must be approximated to $\hat r$ with the help of the information in $D_R$. This is done via datapoints $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ that we then feed to a regressor \cite{vapnik1996nature} to obtain $R^C$, a generalisation of $\hat r$ over the whole state-action space. The complete algorithm is given Alg. \ref{algo:cascading}.
\begin{algorithm}%[H]
    %\small
  \caption{Cascading IRL algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{(s_{j},a_{j},\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}

Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem. Indeed, mere classification only takes into account information from the expert (i.e., which actions goes with what state), but there is more to a MDP than a single policy. Using the Bellman equation in the expression of $\hat r$, we make use of the information lying in the transitions $(s,a,s')$, namely information about the transition probability $P$. The final regression step is just a way to try to generalize this information about $P$ to the whole state-action space, in order to have a well-behaved reward function.

As the analysis (Sec \ref{sec:analysis}) reveals, both classification error and regression error should be controlled. By carefully choosing or tuning both supervised algorithm, it should possible to mitigate any ill effects of the specific settings the cascading algorithm is used in. If one operates in a well-known domain where a lo of data is available, then ease-of-use will probably dictate the choice of two off-the-shelf implementations with default settings. If data is lacking, one could for example have to trade off between bias and variance. The possibility to use and leverage the wide range of techniques developped for these two well known supervised problems is a strong advantage of the proposed cascading approach.
%Reformuler ce passage
The explanation in this section only is a quatitative attempt to make the reader understand the idea behind the algorithm. The analysis of the next section tackle things in a formal, yet less intuitive, way.

@@

In this Section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs. We formulate this result in a more formal way in Theorem \ref{theorem: theorem analysis}

\begin{theorem}
\label{theorem: theorem analysis}
Let $\pi_E$ be the deterministic expert policy, $\rho_E$ its stationary distribution and $\hat{R}^C$ the reward the cascading algorithm outputs. We have:
0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\leq \frac{1}{1-\gamma}(\Delta q \epsilon_C+(1+C_*)\epsilon_R)).
$\Delta q$, and $C_*$ are constants and are defined in the proof. $\epsilon_C$ represents the classification error and $\epsilon_R$ represents the regression error: both terms are well-defined in the proof.
\end{theorem}
\begin{proof}
First let introduce some notations. $q\in\mathbb{R}^{S\times A}$ is the score function outputted by the classification step, $\forall s \in S,\pi_C\in\argmax{a\in A}q(s,a)$ is a deterministic classifier policy and $\epsilon_C$ is the classification error:
\epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
Moreover we recall the definition of the reward $R^C\in\mathbb{R}^{S\times A}$:
\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) +\gamma\sum_{s'\in S}P(s'|s,a)q(\pi_C(s'),a),
and the definition of $\hat{R}^C\in\mathbb{R}^{S\times A}$ which is the reward function outputted by the regression step.
The difference between $R^C$ and $\hat{R}^C$ is noted $\epsilon^R=R^C-\hat{R}^C$, $\epsilon^R\in\mathbb{R}^{S\times A}$ must not be mistaken for the regression error $\epsilon_R\in\mathbb{R}$ defined below.
We also introduce the reward function $R^E\in\mathbb{R}^{S\times A}$ which will be useful in our proof:
\forall (s,a)\in S\times A, R^E(s,a)=q(s,a) +\gamma\sum_{s'\in S}P(s'|s,a)q(\pi_E(s'),a).
Let $X\in\mathbb{R}^{S\times A}$, $\pi\in A^S$, and $a\in A$, we introduce $X_\pi\in\mathbb{R}^S$ and $X_a\in\mathbb{R}^S$ such that: $\forall s\in S, X_\pi(s)=X(s,\pi(s))$ and $\forall s\in S, X_a(s)=X(s,a)$.
Thus we have the following vectorial equations:
\begin{align}
&R^C_a=q_a-\gamma P_aq_{\pi_C},
\\
&R^E_a=q_a+\gamma P_aq_{\pi_E},
\\
&\epsilon^R_a=R^C_a-\hat{R}^C_a.
\end{align}
Now we are going to bound the term: $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]\geq0$ (the lower bound is obvious).
We note $\hat{\pi}_C$ a deterministic optimal policy of the reward $\hat{R}^C$. We have:
V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}=(V^{\hat{\pi}_C}_{\hat{R}^C}-V^{\hat{\pi}_C}_{R^C})+(V^{\hat{\pi}_C}_{R^C}-V^{\pi_E}_{R^C})+(V^{\pi_E}_{R^C}-V^{\pi_E}_{\hat{R}^C}).
We are going to bound each of this three terms. But, first, let $\pi$ be a given deterministic policy. We have:
V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C}=V^{\pi}_{\epsilon^R}=(I-\gamma P_\pi)^{-1}\epsilon^R_{\pi}.
If $\pi=\pi_E$, $\rho_E^T(V^{\pi}_{R^C}-V^{\pi}_{\hat{R}^C})=\rho_E^T(I-\gamma P_{\pi_E})^{-1}\epsilon^R_{\pi_E}=\frac{1}{1-\gamma}\rho_E^T\epsilon^R_{\pi_E}\leq\frac{1}{1-\gamma}\|\epsilon^R_{\pi_E}\|_{1,\rho_E}$.
If $\pi\neq\pi_E$, we introduce the following concentration coefficient:

>>>>

% Created 2012-10-15 lun. 14:49
\documentclass[smallextended]{svjour3}
\smartqed 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subfigure}
\tolerance=1000
\usepackage{amsmath}
%\usepackage{amsthm}
%\newtheorem{definition}{Definition}
%\newtheorem{theorem}{Theorem}
\usepackage{dsfont}
\providecommand{\alert}[1]{\textbf{#1}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
\begin{document}
\title{Learning a reward function from demonstrations: a cascaded supervised learning approach}
\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
\titlerunning{Cascading approach to IRL}
\institute{
 1. ABC Team\\
 LORIA-CNRS, France.\\
\and
2.Supélec\\
 IMS-MaLIS Research group, France\\
 \texttt{firstname.lastname@supelec.fr}\\
\and 
3. UMI 2958 GeorgiaTech-CNRS\\
France\\
}
\date{\today}

@@

\begin{abstract}
  This paper considers the Inverse Reinforcement Learning problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
We propose to break the IRL problem down into two generic Supervised Learning steps: this is the cascaded approach. A classification step followed by a regression step output a reward function.
This approach is justified by an analysis which shows that the demonstrated expert policy is near-optimal for the output reward function. We explain some advantages of this approach over existing algorithms, mainly not needing to repeatedly solve the Reinforcement Learning direct problem. This makes the proposed approach very competitive to the existing algorithms.
Finally, up to the use of some heuristics, we empirically show the algorithm to work with only transitions sampled according to the demonstrated expert policy. We quantitavely demonstrate the soundness of this approach on a highway driving simulator, and illustrate some aspects of the proposed algorithm on the classical inverted pendulum benchmark.
  \end{abstract}
\section{Introduction}
\label{sec-2}

Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} aims at inferring a reward function for which a demonstrated expert policy is optimal. An important part of the literature (see Sec.~\ref{sec:related} for a brief overview) tries to find a reward function such that the associated optimal policy induces a distribution over trajectories (or some measure of this distribution, typically, the feature expectation) starting in a certain state space\footnote{Formally, there is a distribution over the starting state.} which matches the one induced by the expert. Most of these algorithms use Reinforcement
Learning (RL) algorithms as subroutines as finding the optimal policy for a reward function is a necessary step before computing a measure of the previously mentionned distribution. Thus, their performance depends strongly on the quality of these subroutines. Consequently, they suffer from the same challenges of large state spaces, lack of data etc. as RL. In order to avoid repeatedly solving RL problems, we decide to take a different point of view.

Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the inverse of the Bellman equation, see Eq.~\ref{eq:bellman1}), it is worth thinking of a method which is able to output an action-value function for which the greedy policy is the demonstrated expert policy. Thus, the demonstrated expert policy will be optimal for the reward function associated to the action-value function. As a method which is able to output an action-value function for which the greedy policy is the demonstrated expert policy, we propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}). Besides, in order to retrieve, via the inverse of the Bellman optimality equation, the reward associated to the score function the classification outputs, we introduce a regression step (see Sec.~\ref{sec:algo}). That is why the method is called the Cascaded Supervised Approach (CSA). This method is justified by a theoretical analysis (see Sec.~\ref{sec:analysis}) where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.

Finally, an instantiation of CSA is proposed Sec.~\ref{sec:experiments}. This algorithm does not need to iteratively solve a RL problem and needs only sampled transitions from expert and non-experts policies as inputs. Moreover, up to the use of some heuristics, the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy. The algorithm is tested on a car driving simulator and compared to an apprenticeship learning algorithm \cite{abbeel2004apprenticeship}, to a pure classification algorithm \cite{taskar2005learning} and to a recent successfull IRL method \cite{klein2012scirl}.

@@

We work with finite \emph{Markov Decision Processes} (MDPs) \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$. The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamic of the MDP. Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$. At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one optimal deterministic policy.}) policy $\pi\in A^S$. The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$. The reward function $R$ is a local measure of the quality of the control. The global quality of the control induced by a policy $\pi$, with respect to a reward $R$ is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates each state with the discounted cumulative reward for following policy $\pi$ from this state: $V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s]$. Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, V^*_R\geq V^\pi_R$. The stationary distribution induced by a policy $\pi$ is written $\rho_\pi,\textrm{ s.t. }\rho_\pi^TP^\pi = \rho_\pi^T$.

The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by
\begin{equation}
  T^{\pi}_RV = R_\pi + \gamma P_\pi V
  \end{equation}
where $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$ and $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally : $T^*_RV = \max_\pi T^\pi_RV$. Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R: $V^\pi_R$, the fixed point of the Bellman optimality operator $T^*_R$ is the optimal value function $V_R^*$ with respect to reward $R$. Another object object of interest is the action-value function $Q^\pi_R\in\mathbb{R}^{S\times A}$ that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^a_RV = R_a + \gamma P_a V$ with $P_a = (P(s'|s,a))_{s,s' \in S}$) and $R_a$ = $(R(s,a))_{s\in S}$. Once expanded, this gives what we refer to as the Bellman optimality equation :
\begin{equation}
  Q^*_R = R_a + \gamma P_{\pi^*_R}V^*_R
  \label{eq:bellman1}
\end{equation}

An optimal policy follows a greedy mechanism with respect to its optimal action-value function 
\begin{equation}
  \label{eq:greedy}
  \pi^*_R(s)\in\argmax_aQ^*_R(s,a).
\end{equation}

When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, RL or approximate dynamic programming may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}. We call this the direct problem.

In this work, we wish to tackle the inverse problem. We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi^E$, assuming that there exist some unknown reward $R^E$ for which the expert is optimal. The suboptimality of the expert is an interesting setting, that has been discussed for example in \cite{melo2010analysis,syed2012reduction}, but that we are not addressing here due to the numerous complications it introduces. We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is at least near-optimal. The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal). Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see Sec.~\ref{sec:related}.

@@

Our first step in the quest for a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (MC$^2$). This classifier learns a score function $q\in\mathbb{R}^{S\times A}$ that rates the association of a given label $a\in A$ with a certain input $s\in S$. The classification rule $\pi_C\in A^S$ is simply to select (one of) the label(s) that achieves the highest score for the given inputs :

@@

For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as MC$^2$ algorithms, we however used a structured margin approach. We give more examples and detail our choice of a MC$^2$ algorithms for the experiments in Subsec.~\ref{subsec:instanciation}.

Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier. The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL). What is of particular interest to us is the score function $q$ itself. One can easily notice the similarity between Eq.~\ref{eq:greedy2} and Eq.~\ref{eq:greedy} that describes the relation between the optimal policy in a MDP and its optimal action-value function. As the expert is optimal with respect to some unknown reward function $R_E$ and we seek a reward function $R$ for which it is also optimal, it is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$ (which can itself be seen as a good approximation of $\pi_E$). By inversing the Bellman equation, one can express the reward as a function of the action-value function : $R(s,a) = Q^*_R(s,a) - \sum_{s'}P(s'|s,a)Q^*_R(s'\pi^*(s'))$. This leave us with $R^C$, the reward function relative to our score/action-value function $q$ :
\begin{equation}
  R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
\end{equation}
We wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown. For this we use another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitioned to. Action $a_j$ need not be chosen by any particular policy. The dataset $D_R$ brings us information about the dynamics of the system. Our reward function $R^C$ must be approximated with the help of the information in $D_R$. This is done via datapoints $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ that we then feed to a regressor (a simple least-square approximator can do) to obtain $\hat R^C$, a generalisation of $\{(\hat r_j)_j\}$ over the whole state-action space. The complete algorithm is given Alg.~\ref{algo:cascading}.
\begin{algorithm}%[H]
    %\small
  \caption{CSA-IRL algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{(s_{j},a_{j},\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}

Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem. Indeed, mere classification only takes into account information from the expert (i.e., which actions goes with which state), but there is more to a MDP than a single policy. Using the Bellman equation in the expression of $\hat r$, we make use of the information lying in the transitions $(s,a,s')$, namely information about the transition probabilities $P$. The final regression step is just a way to try to generalize this information about $P$ to the whole state-action space, in order to have a well-behaved reward function.

As the analysis (Sec.~\ref{sec:analysis}) reveals, both classification error and regression error should be controlled. By carefully choosing or tuning both supervised learning algorithms, it should be possible to mitigate any ill effects of the specific settings the cascading algorithm is used in. If one operates in a well-known domain where a lot of data is available, then ease-of-use will probably dictate the choice of two off-the-shelf implementations with default settings. If data is lacking, one could for example have to overcome bias, variance or generalization problems in the regression step (more about that in Subsec.\ref{subsec:instanciation}). The possibility to use and leverage the wide range of techniques developped for these two well-known supervised problems is a strong advantage of the proposed cascading approach.

The explanation in this section is only a quatitative attempt to make the reader understand the idea behind the algorithm. The analysis of the next section tackles things in a formal, yet less intuitive, way.

@@

In this section, we prove that the deterministic expert policy $\pi_E$ is near optimal for the reward $\hat{R}^C$ the regression step outputs, more formally, we prove that $\E_{s\sim\rho_E}[V^*_{\hat{R}^C}-V^{\pi_E}_{\hat{R}^C}]$ is small by bounding it by a term that depends on :
\begin{itemize}
  \item the regression and classification errors, respectively $\epsilon_R$ and $\epsilon_C$, these terms are small if the regressor and the classifier are good;
  \item $C_*$, a concentration coefficient assessing the closeness of the stationary distribution of $\pi_E$ and $\hat \pi_C$, the latter being the optimal policy for the reward $\hat R^C$ output by the algorithm ;
  \item $\Delta q=\max_{s\in S}(\max_{a\in A}q(s,a)-\min_{a\in A}q(s,a))=\max_{s\in S}(q(s,\pi_C(s))-\min_{a\in A}q(s,a))$, which could be normalized to $1$ without loss of generality.
  \end{itemize}

\begin{theorem}
\label{theorem: theorem analysis}
Let $\pi_E$ be the deterministic expert policy, $\rho_E$ its stationary distribution and $\hat{R}^C$ the reward the cascading algorithm outputs. We have:
\begin{equation}
0\leq\E_{s\sim\rho_E}[V^*_{\hat{R}^C}(s)-V^{\pi_E}_{\hat{R}^C}(s)]\leq \frac{1}{1-\gamma}\left(\epsilon_C\Delta q +\epsilon_R(1+C_*)\right).
\end{equation}
\end{theorem}
\begin{proof}
First let's introduce some notations. $q\in\mathbb{R}^{S\times A}$ is the score function output by the classification step, $\forall s \in S,\pi_C\in\argmax_{a\in A}q(s,a)$ is a deterministic classifier policy. We recall that $\epsilon_C$ is the classification error:
\begin{equation}
\epsilon_C=\E_{s\sim\rho_E}[\mathds{1}_{\{\pi_C(s)\neq\pi_E(s)\}}].
\end{equation}
Moreover we recall the definition of the reward $R^C\in\mathbb{R}^{S\times A}$:
\begin{equation}
\forall (s,a)\in S\times A, R^C(s,a)=q(s,a) -\gamma\sum_{s'\in S}P(s'|s,a)q(s',\pi_C(s')),
\end{equation}
and the definition of $\hat{R}^C\in\mathbb{R}^{S\times A}$ which is the reward function output by the regression step.
